State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms

### 2023-11-27 manuscript/manuscript_2/paper.tex
Whats with the shift in iota at low SNR?

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman PTA]{Kalman tracking and parameter estimation of continuous gravitational waves with a pulsar timing array: inclusion of pulsar terms}

@@

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is tracked simultaneously with the modulation induced by a single-source gravitational wave in the pulse times of arrival. It is shown how to extend previous state-space analyses to include the pulsar terms. Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the previous approach.
\end{abstract}

@@

\subsection{Parameter estimation} \label{sec:pe_and_ms} 
In this section, we apply the Kalman filter in conjunction with nested sampling to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ for synthetic data $\boldsymbol{Y}$, and compare it to the known, injected values. The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$. That is, given the injected static parameters $\boldsymbol{\theta'}$, multiple realisations of $\boldsymbol{Y}$ are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta'}$. The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations. The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative. Different values of the static GW source parameters are explored in Section \ref{sec:parameter_space}. The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}. All injected static parameters are summarised in the second column of Table \ref{tab:parameters_and_priors}. The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}. \newline

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for \textcolor{red}{YY} arbitrary noise realisations in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. \textcolor{red}{There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours.} The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e. \textcolor{red}{YY} $\%$) out of the \textcolor{red}{YY} possible combinations of the seven parameters and \textcolor{red}{YY} realizations. For this representative example, with characteristic strain \textcolor{red}{XX} the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. \newline

@@

% stacked_GW_plot_small_h_definitive2
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{example-image-a}
	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for nine realisations of the noise processes, with curves coloured uniquely. The horizontal and vertical orange lines indicate the true injected values. The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters. The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$. The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The horizontal axes span a subset of the prior domain for all seven parameters. The known, injected value lies within the 90\% credible interval for \textcolor{red}{X} out of the 63 combinations of seven parameters and nine noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation \textcolor{red}{X} across the 63 combinations.}
	\label{fig:corner_plot_1}
\end{figure*}

%images/corner_lowSNR_chi
	\begin{figure}
	\includegraphics[width=\columnwidth, height =\columnwidth ]{example-image-b}
	\caption{Same as Figure \ref{fig:corner_plot_1}	but for 5 static parameters of $\chi^{(n)}$. The known, injected value lies within the 90\% credible interval for \textcolor{red}{X} out of the \textcolor{red}{X} combinations of five parameters and v\textcolor{red}{X} noise realizations. }
	\label{fig:corner_plot_3}
\end{figure}

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman Pulsar-terms]{State-space analysis of continuous gravitational waves with a pulsar timing array: inclusion of the pulsar terms}

@@

% Abstract of the paper
\begin{abstract}	
	 Pulsar timing arrays may be able to detect continuous nanohertz gravitational waves emitted by individual supermassive black hole binaries. The gravitational wave (GW) signal is composed of two components, the ``Earth term" and ``pulsar term" which correspond to the GW incident on the Earth and the pulsar respectively. The data analysis procedure can be formulated within a state-space framework, wherein intrinsic achromatic spin wandering is Kalman-tracked concurrently with the modulation induced by a single-source gravitational wave in the pulse times of arrival. It is shown how to extend previous state-space analyses to include the pulsar terms. Astrophysically representative software injections in Gaussian measurement noise evidence that the inclusion of the pulsar terms eliminates the biases in the parameter estimates, increasing the parameter accuracy by \textcolor{red}{X percent} and the minimal detectable strain by \textcolor{red}{Y percent} relative to the previous approach.
\end{abstract}

@@

\subsection{Parameter estimation} \label{sec:pe_and_ms} 
In this section, we apply the Kalman filter in conjunction with nested sampling to calculate the joint posterior probability distribution $p({\boldsymbol{\theta'}} | {\boldsymbol{Y}})$ for synthetic data $\boldsymbol{Y}$, and compare it to the known, injected values. The procedure is undertaken for multiple realisations of the noise processes $\xi^{(n)}(t)$ and $\varepsilon^{(n)}(t)$. That is, given the injected static parameters $\boldsymbol{\theta'}$, multiple realisations of $\boldsymbol{Y}$ are generated using the methods described in Section \ref{sec:synthetic_data}, and for each realisation we independently estimate $\boldsymbol{\theta'}$. The aims are (i) to demonstrate that the analysis scheme works (i.e.\ that it converges to a well-behaved, unimodal posterior for multiple noise realisations), (ii) to give a preliminary sense of its accuracy, and (iii) quantify the natural random dispersion in the one-dimensional posterior medians between noise realisations. The injected static GW source parameters $\boldsymbol{\theta}_{\rm gw}$ used for this mock analysis to generate $\boldsymbol{Y}$ are selected to be astrophysically reasonable and representative. Different values of the static GW source parameters are explored in Section \ref{sec:parameter_space}. The static pulsar parameters $\boldsymbol{\theta'}_{\rm psr}$  are as specified in Section \ref{sec:synthetic_data}. All injected static parameters $\boldsymbol{\theta}'$ are summarised in the second column of Table \ref{tab:parameters_and_priors}. The specification of the priors on $\boldsymbol{\theta'}$ is described in Appendix \ref{sec:set_priors} and summarised in the rightmost column of Table \ref{tab:parameters_and_priors}. \newline

Figure \ref{fig:corner_plot_1} displays results for the seven parameters in $\boldsymbol{\theta}_{\rm gw}$ for 10 arbitrary noise realisations in the form of a traditional corner plot. The histograms are the one-dimensional posteriors for each parameter, marginalized over the six other parameters. Each coloured curve is the result for different noise realisation. The solid orange line marks the known injected value. The two-dimensional contours mark the (0.5, 1, 1.5, 2)-sigma level surfaces. All histograms and contours are consistent with a unimodal joint posterior, which peaks near the known, injected values. There is scant evidence of railing against the prior bounds. There is no strong evidence for correlations between parameter pairs, e.g.\ banana-shaped contours. The injected value is contained within the 90\% credible interval of the one-dimensional marginalized posteriors in \textcolor{red}{YY} (i.e. \textcolor{red}{YY} $\%$) out of the \textcolor{red}{YY} possible combinations of the seven parameters and \textcolor{red}{YY} realizations. For this representative example, with characteristic strain \textcolor{red}{XX} the analysis scheme estimates $\boldsymbol{\theta}_{\rm gw}$ accurately. \newline

@@

% stacked_GW_plot_small_h_definitive2
\begin{figure*}
	\includegraphics[width=\textwidth, height =\textwidth ]{images/corner_lowSNR}
	\caption{Posterior distribution of the GW source parameters $\boldsymbol{\theta}_{\rm gw}$ for the representative system in Table \ref{tab:parameters_and_priors}, for nine realisations of the noise processes, with curves coloured uniquely. The horizontal and vertical orange lines indicate the true injected values. The contours in the two-dimensional histograms mark the (0.5, 1, 1.5, 2)-$\sigma$ levels after marginalizing over all but two parameters. The one-dimensional histograms correspond to the joint posterior distribution marginalized over all but one parameter. The supertitles of the one-dimensional histograms record the posterior median and the 0.16 and 0.84 quantiles of the median realisation. We plot the scaled variables $10^9 \Omega$ (units: ${\rm rad \, s^{-1}}$) and $10^{15} h_0$. The Kalman filter and nested sampler estimate accurately all seven parameters in ${\boldsymbol{\theta}}_{\rm gw}$. The horizontal axes span a subset of the prior domain for all seven parameters. The known, injected value lies within the 90\% credible interval for \textcolor{red}{X} out of the 63 combinations of seven parameters and nine noise realizations. There is an appreciable dispersion among the peaks of the one-dimensional posteriors, with coefficient of variation \textcolor{red}{X} across the 63 combinations.}
	\label{fig:corner_plot_1}
\end{figure*}

%images/corner_lowSNR_chiF
	\begin{figure}
	\includegraphics[width=\columnwidth, height =\columnwidth ]{images/corner_lowSNR_chi}
	\caption{Same as Figure \ref{fig:corner_plot_1}	but for 5 static parameters of $\chi^{(n)}$. The known, injected value lies within the 90\% credible interval for \textcolor{red}{X} out of the \textcolor{red}{X} combinations of five parameters and v\textcolor{red}{X} noise realizations. }
	\label{fig:corner_plot_3}
\end{figure}
